Long- and Short-Read Metabarcoding Give Similar Results for Spatial Turnover Range in an Ectomycorrhizal Savanna Woodland

### 2020-02-13 writing/transect_paper.Rmd
Complete draft as sent to Anna

---
title: "A Tale of Two Transects"
subtitle: "Also Two Sequencing Technologies, and Two Amplicons"
author: 
  - Brendan Furneaux^1^,
    Mohammad Bahram^1^,
    Anna Rosling^3^,
    Nourou S. Yorou^2^,
    Martin Ryberg^1^
  - ^1^Program in Systematic Biology,
    Department of Organismal Biology,
    Uppsala University,
    Uppsala, Sweden
  - ^2^Research Unit in Tropical Mycology and Plant-Fungi Interactions,
    LEB,
    University of Parakou,
    Parakou, Benin
  - ^3^Program in Evolutionary Biology,
    Department of Ecology and Genetics,
    Uppsala University,
    Uppsala, Sweden
date: "`r format(Sys.time(), '%B %d, %Y')`"
bibliography:
  - "all.bib"
  - "R.bib"
csl: molecular-ecology.csl
always_allow_html: true
output:
  bookdown::word_document2: 
    keep_md: yes
  bookdown::pdf_document2:
    fig_crop: yes
    keep_tex: yes
    latex_engine: lualatex
    includes:
      in_header: preamble.tex
---

@@

@@

## Introduction to ECM communities

Ectomycorrhizas (ECM) are a symbiosis between fungi and plant root tips.
The fungi provide nutrients and protection from pathogens in exchange for sugards from the plant [@Smith2010].
Approximately 8% of described fungal species [@ainsworth2008ainsworth;@Rinaldi2008] and 2% of land plant species [@brundrett2017a] form ECM.
The plants include ecologically and economically important stand-forming trees belonging to both temperate/boreal groups such as Pinaceae and Fagaceae, and tropical groups such as Dipterocarpaceae, and Fabaceae tr. Amherstieae.

Although many well-known mushrooms are produced by ECM fungi (e.g., *Amanita*, *Cantharellus*, *Boletus*), many ECM fungi instead produce inconspicuous (e.g., *Tomentella*) or no (e.g., *Cenococcum*) fruit bodies.
Even when fruitbodies are large, they are ephemeral, so study of ECM communities is facilitated by looking at vegetative structures.
Unlike many saprotrophic fungi which grow easily in axenic culture, ECM fungi are usually difficult to culture, so DNA barcoding is increasingly used to investigate vegetative structures.
The advent of high-thoughput sequencing (HTS) has facilitated such studies by providing enough sequencing depth for metabarcoding of bulk environmental samples such as soils.

## Metabarcoding methods

As additional techniques and methods are developed for HTS, there is an increasing array of choices for researchers investigating fungal communities.
Traditionally, fungal metabarcoding studies have targetted the ITS1 or ITS2 regions, which provide sufficient resolution to distinguish fungal species in many groups, and which are usually short enough for HTS [@schoch2012;@Lindahl2013].
The resulting sequencing reads are clustered by sequence similarity to form operational taxonomic units (OTUs), which are then used as the units for further community analysis.
If taxonomic identification is desired, it has usually been performed by database searches using BLAST (ref).
This approach comes with some weaknesses.

While ITS1 and ITS2 often have suitable variation to distinguish species, they cannot be reliably aligned over the large taxonomic scales typically present in environmental samples.
Additionally the wide range of length variation of these regions may introduce bias in recovery of different taxa.
Further bias is introduced by variation in the 5.8S region which separates the two ITS regions, which makes it difficult to design primers that are suitable for all fungi [@tedersoo2015].
The recent advent of long-read HTS technologies such as Pacific Biosciences Single Molecule Real Time sequencing (PacBio) introduce the possibility of sequencing longer amplicons which include both the ITS regions and the flanking, more conserved SSU and/or LSU regions [@Tedersoo2018].
This has the potential to improve taxonomic placement of sequences without close database matches, and to allow the alignment of metabarcoding reads.
However, long-read technologies are currently more expensive per read, and so their use entails a trade-off with sequencing depth and/or sample number [@Kennedy2018].

Distance-based clustering conflates intra-species variation and sequencing error, and results are dataset-specific.
In contrast, more recent denoising methods such as DADA2 [@callahan2017], Deblur [@amir2017], and UNOISE2 [@edgar2016] utilize read quality information to control for sequencing error while leaving intra-species variation.
The resulting units are known as amplicon sequence variants (ASVs) or exact sequence variants (ESVs), as they should represent all true amplicon sequences from the sample.
Unlike cluster-based OTUs, ASVs can capture variation of as little as one base pair, and are not as dataset specific [@callahan2017].

@@

Phylogenetic information has the potential to both improve taxonomic assignment and provide alternative measures of community alpha and beta diversity.
Because OTU clustering may both "clump" different species in a single OTU, and "split" a single species into multiple OTUS [@ryberg2015], diversity measures based on counting species within a community or shared species between two communities may give different results depending on the clustering threshold.
In contrast, phylogenetic community distance measures like UNIFRAC [@wong2016] are relatively insensitive to species/OTU delimitation, but require a phylogenetic tree.
Phylogenetic placement algorithms [@matsen2010; @berger2011] are not easy to apply to ITS sequences because they require that the query sequences be aligned to a reference alignment.
Additionally, methods exist to place OTUs on a simplified tree based on taxonomic assignments [@tedersoo2018].
However, long amplicon sequencing allows the inclusion of alignable regions for construction of more fully resolved phylogenetic trees directly from metabarcoding reads.

@@

## Turnover scale in ECM forests

@@

An additional sample was collected at every sampling location (1-m spacing) in 2016 using LifeGuard^TM^ Soil Preservation Solution (MO BIO, Carlsbad, CA; USA) for preservation, without field lysis.
Sequencing results for these samples were found not to be comparable with results preserved using the Xpedition^TM^ lysis buffer (Supplementary figure \@ref(fig:)), so these samples were not included in spatial analyses.
However, reads from these sequences were included in the full bioinformatics workflow, including ASV calling, OTU clustering, and phylogenetic trees.

@@

Two different fragments of the nuclear rDNA were amplified (Figure \@ref(fig:rDNA)).
The short amplicon (approximately 300 bp) included the full ITS2 region as well as parts of the flanking 5.8S and large subunit (LSU) rDNA, using gITS7 [@ihrmark2012] as the forward primer and a mix of ITS4 [@white1990amplification] and ITS4a [@urbina2016] as the reverse primer (hereafter, ITS4m).
The long amplicon (approximately 1500 bp) included approximately 12 bp at the 3' end of the small subunit (SSU) rDNA, the full ITS region including the 5.8S rDNA, and approximately 950 bp at the 5' end of the LSU, including the first three variable regions (Figure \@ref(fig:rDNA)), using ITS1 [@white1990amplification] as the forward primer and LR5 [@vilgalys1990] as the reverse primer.

The gITS7 primers for the short amplicon were indexed for multiplexing (**supp info**).
Amplification was performed by polymerase chain reaction (PCR) in 20µl reactions containing 200 µM dNTP mix, 250 µM indexed gITS7 primer, 150µM ITS4m, 2mM McCl_2_, 0.1 U *Taq* polymerase (Dream *Taq*, Thermo Fisher Scientific, Waltham, MA, USA) and 3--7 ng purified DNA in Dream *Taq* buffer.
The reaction conditions were 10 min at 95°, followed by 35 cycles of 60 s at 95°, 45 s at 56°, and 50 s at 72°, and finally 3 min at 72°.
Each reaction was conducted in three technical replicates to reduce the effect of PCR stochasticity, which were pooled after amplification.

@@

Both primers for the long amplicon were indexed for multiplexing (**supp info**).
PCR was performed as for the short amplicons, but with 500 µM of each of the two primers.
Reaction conditions were 10 min at 95°, 30 cycles of 45 s at 95°, 45 s at 59°, and 90 s at 72°, and finally 10 min at 72°.
Each reaction was performed in three technical replicates as for short amplicons.

@@

Amplicons were purified using SPRI beads [@vesterinen2016] and quantified fluorometrically as above.
An aliquot of 100 ng of DNA from each sample (or the total PCR product if less than 100 ng) was pooled into two libraries each for long and short amplicons.
The first library contained all samples from 2016 plus one PCR blank and one positive control (68 total samples), and the second containes all samples from 2015 plus one PCR blank and one positive control (42 total samples).
Each library was sequenced using Single Molecule Real Time (SMRT) sequencing on a Pacific Biosciences (PacBio) RS II sequencer at the Uppsala Genome Center (UGC; Uppsala Genome Center, Science
for Life Laboratory, Dept. of Immunology, Genetics and Pathology, Uppsala University, BMC, Box 815, SE-752 37 UPPSALA, Sweden).
Short amplicon libraries were sequenced on two SMRT cells each, while long amplicon libraries were sequenced on four SMRT cells each.

@@

Forward and reverse Illumina reads were truncated after the first base with quality score <= 10, and then reads with more than 3 expected errors in either read were discarded.
Forward and reverse reads were denoised using `DADA2` version `r packageVersion("dada2")` [@callahan2016] using seperate error models and then merged.

@@

Raw reads were divided into shorter regions by matching to covariance models (CM), which are similar to stochastic hidden markov models (HMM), but account for both nucleotide sequence and RNA secondary structure [@eddy1994].
First, the 5.8S rDNA was located in each read by searching for Rfam model RF0002 [@kalvari2018] using `cmsearch` from Infernal 1.1.2 [@nawrocki2013], and all bases before the 5.8S were assigned to ITS1.
No attempt was made to remove the approximately 12 bp fragment of the SSU from the 5' end of ITS1 in the long amplicons; it was too short to be reliably detected by a CM or the HMMs employed by ITSx [@bengtsson-palme2013].
<!--This approach was able to identify both the 5' and 3' ends of the 5.8S rDNA more consistently and quickly than ITSx [@bengtsson-palme2013, data not shown].-->
A reference alignment including conserved RNA base pairing between and within the 5.8S and relevant portions of LSU was generated from the fungal 28S RNA seed alignment from the Ribosomal Data Project (RDP) release 11.5 [@glockner2017; @cole2014] by truncating after the LR5 primer site and using the reference line to annotate the variable regions *sensu* @michot1984 and @raue1988.
A CM was generated from the alignment using `cmbuild` from Infernal.
The fragment of each read beginning with the 5.8S rDNA was then aligned to the CM using `cmalign` from Infernal.
The annotation line in the CM alignment for each read was then used to split the reads into alternating more-conserved and less-conserved regions as shown in Figure \@ref(fig:rDNA),
where LSU1-4 represent the conserved regions of LSU flanking the variable D1-3 regions [@michot1984].
For short amplicons, only (partial) 5.8S, ITS2, and (partial) LSU1 were extracted.
Code to extract the regions, including annotated seed alignments and CMs, is available in the new R package `LSUx`, available on github (https://github.com/brendanf/LSUx).

Each of the extracted regions was independently filtered for length (Table \@ref(tab:region-length)) and a maximum of three expected errors.
Sequences were then dereplicated and denoised into amplicon sequencing variants (ASVs) using `DADA2` version `r packageVersion("dada2")` [@callahan2016; @callahan2019].
The error model for `DADA2` denoising was fit using the 5.8S region for long amplicons, and using the entire read for short amplicons.
Independent error models were fit for each sequencing run (i.e., long *vs.* short amplicons, different sequencing technologies).
For PacBio libraries, `DADA2` was run with complete pooling and a band size of 16.
For Ion Torrent libraries, pseudo-pooling and a band size of 32 were used, and the homopolymer gap penalty was set to -1, as recommended by the `DADA2` FAQ (https://benjjneb.github.io/dada2/faq.html).
Chimeras within each region were removed using `removeBimeraDenovoTable` from `DADA2`.

For each ITS2 ASV from the long amplicon data set, the denoised sequences for the other regions corresponding to the same sequencing reads were concatenated to form a set of full-length reads.
For reads which were not assigned a denoised sequence for each region, the raw read for the region was used instead.
Because ITS2 is the most variable of the amplified regions (Supplementary Figure \@ref(fig:regions)), reads with identical ITS2 regions are expected have highly similar sequences in the other regions, unless the amplicon was chimeric.
The concatenated ASVs representing each long read were aligned in R using the `DECIPHER` package [@wright2015].
Outlier sequences, as determined by mean pairwise distance from the rest of the alignment, were removed from each alignment using the `odseq` package [@jehl2015], using the default threshold of 0.025.
The consensus of the remaining aligned sequences was assigned as the full-length ASV sequence.
Full-length ASV sequences with more than three ambiguous bases (i.e., no nucleotide >50% at a given position) were removed *(how many)*.
The count and sample distribution of reads assigned to each ASV were calculated in order to form a sample x ASV community matrix.
A similar process was used to generate a consensus ITS (ITS1--5.8S--ITS2) and LSU (LSU1--D1--LSU2--D2--LSU3--D3--LSU4) sequence for each ASV.
The process of assigning consensus full-length ASVs was carried out using the new `tzara` package for R, available on github (https://github.com/brendanf/tzara).

@@

Unique LSU regions from long amplicon ASVs were aligned using DECIPHER [@wright2015] with up to 10 iterations of progressive alignment and conserved RNA secondary structure calculation and 10 refinement iterations.
<!-- The LSU tree ends up not looking as nice as the tree which includes ITS (I don't know why, except that maybe 5.8S is more helpful than ITS is harmful?) so I'm using that. Maybe I don't need to mention the LSU-only tree.-->
This alignment was truncated at a position after the D3 region corresponding to base 907 of the *Saccharomyces cerevisiae* S288C reference sequence for LSU, because several sequences had type 1 introns
<!-- Just curious, what are Type 1 introns -AR--> <!-- It has to do with the enzyme that cuts them out, I think, but I don't actually know my introns are Type 1, just that this is a location where Type 1 introns are known to occur. -BF-->
after this position [@holst-jensen1999].
Full length long amplicon ASVs (including ITS1, 5.8S, and ITS2 regions) were aligned and truncated in the same way.

@@

Taxonomic assignments of ASVs from the long amplicon dataset were refined using the phylogenetic tree (Supplementary Figure \@ref(fig:phylotax)).
A taxon at a particular rank was assigned to a node and all its descendents if that taxon was consistent with the reference-based taxonomic assignments for each of the descendents.
A taxon assignment was considered to be consistent if at least one algorithm assigned that taxon at greater than 50% confidence, or if no algorithm successfully classified the sequence at greater than 50% confidence.
The result of this process was to give a taxonomic assignment to ASVs which were previously unassigned if they were nested within a clade where which was consistently given an assignment, as well as clarify the assignment of ASVs where different algorithms had resulted in different assignments, but only one of these was consistent with the assignments of other ASVs in the same clade.
This refinement algorithm is referred to as "PHYLOTAX".

@@

Additionally, empirical distance-decay curves were generated by plotting mean community dissimilarity as a function of spatial distance, and fit to an exponential model of the form given by @Legendre2012 using the `nls` function in R.
Points in the empirical distance-decay curve were weighted by the number of comparisons within the distance class and the inverse of the distance for the purposes of model fitting.
For datasets where the Mantel corellogram indicated spatial correlation between samples taken in seperate years, the model was re-fit with an additional term to represent temporal correlation:
$$D = C_0 + C_1\left[1 - \exp\left(-3  \left(\frac{d}{a_d} + \frac{t}{a_t}\right)\right)\right]$$
where $D$, $d$, and $t$ represent the community dissimilarity, spatial distance, and time lag between samples, respectively, and the parameters are $C_0$, the community dissimilarity from replicate samples ("nugget"); $C_0 + C_1$, the community dissimilarity at long distances ("sill"); $a_d$ the spatial range at which the community dissimilarity has moved 95% of the way from "nugget" to "sill", and $a_t$ the equivalent temporal range.
95% confidence intervals were calculated for the spatial and temporal range parameters by profiling using the `MASS` package in R.

@@

The number of sequencing reads and ASVs at each stage in the bioinformatics pipeline are shown in Table \@ref(tab:bioinfo).
Long (short) amplicon sequencing with PacBio yielded `r readd(rawcounts)["pb_500"]` (`r readd(rawcounts)["pb_483"]`) circular consensus reads, of which `r readd(demuxcounts)["pb_500"]` (`r readd(demuxcounts)["pb_483"]`) were successfully trimmed and demultiplexed.
Illumina sequencing yielded `r readd(rawcounts)["SH-2257"]` read pairs, of which `r readd(demuxcounts)["SH-2257"]` were successfully trimmed and demultiplexed, and `r readd(filtercounts_full)["SH-2257"]` passed quality filtering.
Ion Torrent sequencing yielded `r readd(demuxcounts)["is_057"]` demultiplexed reads.

@@

### Region extraction

- Fraction where 5.8S was detected
- Length/ASV count/ASV fraction of extracted regions.

### Reproducibility of read counts from different technologies

(ref:venn) Venn diagrams showing shared ITS2-based ASVs (left) and 97% OTUs (right) between different sequencing strategies.
In each region, the number of ASVs/OTUs is given above, while the fractions of reads for each sequencing strategy are shown below.

```{r venn, fig.cap = "(ref:venn)", out.height = "30%", fig.show="hold"}
readd(vennplot_ASV) %>% {
  plot(
    select(., set, ASVs) %>% deframe() %>% replace_na(0) %>% eulerr::venn(),
    legend = list(
      labels = c("PacBio Long", "PacBio Short", "Ion Torrent", "Illumina")
    ),
    quantities = .$label,
    main = "ASV"
  )
}

readd(vennplot_OTU) %>% {
  plot(
    select(., set, OTUs) %>% deframe() %>% replace_na(0) %>% eulerr::venn(),
    legend = FALSE,
    labels = FALSE,
    quantities = .$label,
    main = "OTU"
  )
}
```

(ref:venn2) (*alternate to \@ref(fig:venn)*) Venn diagrams showing shared ITS2-based ASVs (left) and 97% OTUs (right) between different sequencing technologies from the same short amplicon library.
In each region, the number of ASVs/OTUs is given above, while the fractions of reads for each sequencing strategy are shown below.

>>>>

---
title: "Long- and Short-Read Metabarcoding Give Similar Results for Spatial Turnover Range in an Ectomycorrhizal Savanna Woodland"
author: 
  - Brendan Furneaux^1^,
    Mohammad Bahram^1^,
    Anna Rosling^3^,
    Nourou S. Yorou^2^,
    Martin Ryberg^1^
  - ^1^Program in Systematic Biology,
    Department of Organismal Biology,
    Uppsala University,
    Uppsala, Sweden
  - ^2^Research Unit in Tropical Mycology and Plant-Fungi Interactions,
    LEB,
    University of Parakou,
    Parakou, Benin
  - ^3^Program in Evolutionary Biology,
    Department of Ecology and Genetics,
    Uppsala University,
    Uppsala, Sweden
date: "`r format(Sys.time(), '%B %d, %Y')`"
bibliography:
  - "all.bib"
  - "R.bib"
csl: molecular-ecology.csl
always_allow_html: true
output:
  bookdown::word_document2: 
    keep_md: yes
  bookdown::pdf_document2:
    fig_crop: yes
    keep_tex: yes
    latex_engine: lualatex
    includes:
      in_header: preamble.tex
---

@@

# Abstract

Ectomycorrhizal (ECM) communities are characterized by mutualistic symbiosis between certain fungi and plant roots.
Spatial turnover rate is one important characteristic of ecological communities, which is relevant for biodiversity assessment and sampling design.
We investigated turnover rate in an ECM fungal community using soil metabarcoding with four different sequencing strategies: short amplicon sequencing of the ITS2 region (300--400 bp) with Illumina MiSeq, Ion Torrent Ion S5, and PacBio RS II, as well as long amplicon sequencing of the full ITS and partial LSU regions (1200--1600 bp) with PacBio RS II.
Resulting turnover rates depended more on statistical method than sequencing technology.
We found significant autocorrelation in distance classes <3 m using a Mantel correlogram, and a best-fit distance-decay curve range of 17--19 m, both shorter than previously reported for similar habitats.
The use of long-amplicon sequencing enables construction of a phylogenetic tree from metabarcoding reads, which facilitates taxonomic identification of sequences.
However, long reads present issues for denoising algorithms in diverse communities.
We present a solution that splits the reads into shorter homologous regions prior to denoising, and then reconstructs the full denoised reads.
In the choice between short and long amplicons, we suggest a hybrid approach using short amplicons for sampling breadth and depth, and long amplicons to characterize the local species pool for improved identification and phylogenetic analyses.

@@

Ectomycorrhizas (ECM) are a symbiosis between fungi and plant root tips.
The fungi provide nutrients and protection from pathogens in exchange for sugars from the plant [@Smith2010].
Approximately 8% of described fungal species are believed to take part in ECM symbiosis [@ainsworth2008ainsworth;@Rinaldi2008].
Although only about 2% of land plant species form ECM, these include ecologically and economically important stand-forming trees belonging to both temperate/boreal groups such as Pinaceae and Fagaceae, and tropical groups such as Dipterocarpaceae, and Fabaceae tr. Amherstieae [@brundrett2017a].

Although many well-known mushrooms are produced by ECM fungi (e.g., *Amanita*, *Cantharellus*, *Boletus*), many ECM fungi instead produce inconspicuous (e.g., *Tomentella*) or no (e.g., *Cenococcum*) fruit bodies.
Even when fruitbodies are large, they are ephemeral, so study of ECM communities is facilitated by looking at vegetative structures [@horton2001].
Unlike many saprotrophic fungi which grow easily in axenic culture, ECM fungi are usually difficult to culture, so DNA barcoding is increasingly used to investigate vegetative structures.
The advent of high-thoughput sequencing (HTS) has facilitated such studies by providing enough sequencing depth for metabarcoding of bulk environmental samples such as soils [@Lindahl2013].

As additional techniques and methods are developed for HTS, there is an increasing array of choices for researchers investigating fungal communities.
Traditionally, fungal metabarcoding studies have targetted the ITS1 or ITS2 regions, which provide sufficient resolution to distinguish fungal species in many groups, and which are usually short enough for HTS [@schoch2012;@Lindahl2013].
The resulting sequencing reads are clustered by sequence similarity to form operational taxonomic units (OTUs), which are then used as the units for further community analysis [@Lindahl2013].
If taxonomic identification is desired in order to put OTUs in a wider context and associate functional information, it has usually been performed by database searches using BLAST [@altschul1990;@Lindahl2013].
However, this traditional approach comes with some potential weaknesses.

While ITS1 and ITS2 often have suitable variation to distinguish species, they cannot be reliably aligned over the large taxonomic scales typically present in environmental samples [@Lindahl2013].
Additionally the wide range of length variation of these regions may introduce bias in recovery of different taxa.
Further bias is introduced by variation in the 5.8S region which separates the two ITS regions, which makes it difficult to design primers that are suitable for all fungi [@tedersoo2015].
The recent advent of long-read HTS technologies such as Pacific Biosciences Single Molecule Real Time sequencing (PacBio) introduce the possibility of sequencing longer amplicons which include both the ITS regions and the flanking, more conserved SSU and/or LSU regions [@Tedersoo2018].
This has the potential to improve taxonomic placement of sequences that lack close database matches, and to allow the alignment of metabarcoding reads.
However, long-read technologies are currently more expensive per read, and so their use entails a trade-off with sequencing depth and/or sample number [@Kennedy2018].

Distance-based clustering conflates intra-species variation and sequencing error, and results are dataset-specific.
In contrast, more recent denoising methods such as DADA2 [@callahan2017], Deblur [@amir2017], and UNOISE2 [@edgar2016] utilize read quality information to control for sequencing error while leaving intra-species variation.
The resulting units are known as amplicon sequence variants (ASVs) or exact sequence variants (ESVs), as they should represent true amplicon sequences from the sample.
Unlike cluster-based OTUs, ASVs can capture variation of as little as one base pair, and are not as dataset specific [@callahan2017].

@@

Phylogenetic information has the potential to both improve taxonomic assignment and provide alternative measures of community alpha and beta diversity.
Because OTU clustering may both "clump" different species into a single OTU, and "split" a single species into multiple OTUs [@ryberg2015], diversity measures based on counting species within a community or shared species between two communities may give different results depending on the clustering threshold.
In contrast, phylogenetic community distance measures like UNIFRAC [@wong2016] are relatively insensitive to species/OTU delimitation, but require a phylogenetic tree.
Phylogenetic placement algorithms have been developed to place short amplicon reads onto a reference tree [@matsen2010; @berger2011], but are not easy to apply to ITS sequences because they require that the query sequences be aligned to a reference alignment.
Additionally, methods exist to place OTUs on a simplified tree based on taxonomic assignments [@tedersoo2018].
However, long amplicon sequencing allows the inclusion of alignable regions for construction of more fully resolved phylogenetic trees directly from metabarcoding reads.

@@

@@

An additional sample was collected at every sampling location (1-m spacing) in 2016 using LifeGuard^TM^ Soil Preservation Solution (MO BIO, Carlsbad, CA; USA) for preservation, without field lysis.
Sequencing results for these samples were found not to be comparable with results preserved using the Xpedition^TM^ lysis buffer (Supplementary Figures \@ref(fig:buffer-compare-long) and \@ref(fig:buffer-compare-short)), so these samples were not included in spatial analyses.
However, reads from these sequences were included in the full bioinformatics workflow, including ASV calling, OTU clustering, and phylogenetic trees.

@@

Two different fragments of the nuclear rDNA were amplified (Figure\ \@ref(fig:rDNA)).
The short amplicon (approximately 300 bp) targetted the full ITS2 region as well as parts of the flanking 5.8S and large subunit (LSU) rDNA, using gITS7 [@ihrmark2012] as the forward primer and a mix of ITS4 [@white1990amplification] and ITS4a [@urbina2016] as the reverse primer (hereafter, ITS4m).
The long amplicon (approximately 1500 bp) targetted the full ITS region including the 5.8S rDNA and approximately 950 bp at the 5' end of the LSU, including the first three variable regions (Figure\ \@ref(fig:rDNA)), using ITS1 [@white1990amplification] as the forward primer and LR5 [@vilgalys1990] as the reverse primer.

The gITS7 primers for the short amplicon were indexed for multiplexing (**Supplementary File**).
Amplification was performed by polymerase chain reaction (PCR) in 20µl reactions containing 200 µM dNTP mix, 250 µM indexed gITS7 primer, 150µM ITS4m, 2mM MgCl~2~, 0.1 U *Taq* polymerase (Dream *Taq*, Thermo Fisher Scientific, Waltham, MA, USA) and 3--7 ng purified DNA in Dream *Taq* buffer.
The reaction conditions were 10 min at 95°, followed by 35 cycles of 60 s at 95°, 45 s at 56°, and 50 s at 72°, and finally 3 min at 72°.
Each reaction was conducted in three technical replicates to reduce the effect of PCR stochasticity, which were pooled after amplification.

@@

Both primers for the long amplicon were indexed for multiplexing (**Supplementary File**).
PCR was performed as for the short amplicons, but with 500 µM of each of the two primers.
Reaction conditions were 10 min at 95°, 30 cycles of 45 s at 95°, 45 s at 59°, and 90 s at 72°, and finally 10 min at 72°.
Each reaction was performed in three technical replicates as for short amplicons.

@@

Amplicons were purified using SPRI beads [@vesterinen2016] and quantified fluorometrically as above.
An aliquot of 100 ng of DNA from each sample (or the total PCR product if less than 100 ng) was pooled into two libraries each for long and short amplicons.
Each library was sequenced using Single Molecule Real Time (SMRT) sequencing on a Pacific Biosciences (PacBio) RS II sequencer at the Uppsala Genome Center (UGC; Uppsala Genome Center, Science
for Life Laboratory, Dept. of Immunology, Genetics and Pathology, Uppsala University, BMC, Box 815, SE-752 37 UPPSALA, Sweden).
Short amplicon libraries were sequenced on two SMRT cells each, while long amplicon libraries were sequenced on four SMRT cells each.

@@

Forward and reverse Illumina reads were truncated after the first base with quality score <= 10, and then reads with more than 3 expected errors in either read were discarded.
Forward and reverse reads were denoised using `DADA2` version `r packageVersion("dada2")` [@callahan2016] using seperate error models and pseudo-pooling, and then forward and reverse reads were merged.

@@

Raw reads were divided into shorter regions by matching to covariance models (CM), which are similar to stochastic hidden markov models (HMM), but account for both nucleotide sequence and RNA secondary structure [@eddy1994].
First, the 5.8S rDNA was located in each read by searching for Rfam model RF0002 [@kalvari2018] using `cmsearch` from Infernal 1.1.2 [@nawrocki2013], and all bases before the 5.8S were assigned to ITS1.
No attempt was made to remove the approximately 12 bp fragment of the SSU from the 5' end of ITS1 in the long amplicons; it was too short to be reliably detected by a CM or the HMMs employed by ITSx [@bengtsson-palme2013].
<!--This approach was able to identify both the 5' and 3' ends of the 5.8S rDNA more consistently and quickly than ITSx [@bengtsson-palme2013, data not shown].-->
A reference alignment including conserved RNA base pairing between and within the 5.8S and relevant portions of LSU was generated from the fungal 28S RNA seed alignment from the Ribosomal Data Project (RDP) release 11.5 [@glockner2017; @cole2014] by truncating after the LR5 primer site and using the reference line to annotate the variable regions *sensu* @michot1984 and @raue1988.
A CM was generated from the alignment using `cmbuild` from Infernal.
The fragment of each read beginning with the 5.8S rDNA was then aligned to the CM using `cmalign` from Infernal.
The annotation line in the CM alignment for each read was then used to split the reads into alternating more-conserved and less-conserved regions as shown in Figure\ \@ref(fig:rDNA),
where LSU1-4 represent the conserved regions of LSU flanking the variable D1-3 regions [@michot1984].
For short amplicons, only (partial) 5.8S, ITS2, and (partial) LSU1 were extracted.
Code to extract the regions, including annotated seed alignments and CMs, is available in the new R package `LSUx`, available on github (https://github.com/brendanf/LSUx).

Each of the extracted regions was independently filtered for length (Supplementary Table\ \@ref(tab:region-limits)) and a maximum of three expected errors.
Sequences were then dereplicated and denoised into amplicon sequencing variants (ASVs) using `DADA2` version `r packageVersion("dada2")` [@callahan2016; @callahan2019].
The error model for `DADA2` denoising was fit using the 5.8S region for long amplicons, and using the entire read for short amplicons.
Independent error models were fit for each sequencing run (i.e., long *vs.* short amplicons, different sequencing technologies).
For PacBio libraries, `DADA2` was run with complete pooling and a band size of 16.
For Ion Torrent libraries, pseudo-pooling and a band size of 32 were used, and the homopolymer gap penalty was set to -1, as recommended by the `DADA2` FAQ (https://benjjneb.github.io/dada2/faq.html).
Chimeras within each region were removed using `removeBimeraDenovoTable` from `DADA2`.

For each ITS2 ASV from the long amplicon data set, the denoised sequences for the other regions corresponding to the same sequencing reads were concatenated to form a set of full-length reads.
For reads which were not assigned a denoised sequence for each region, the raw read for the region was used instead.
Because ITS2 is the most variable of the amplified regions (Supplementary Figure\ \@ref(fig:regions)), reads with identical ITS2 regions are expected have highly similar sequences in the other regions, unless the amplicon was chimeric.
The concatenated ASVs representing each long read were aligned in R using the `DECIPHER` package [@wright2015].
Outlier sequences, as determined by mean pairwise distance from the rest of the alignment, were removed from each alignment using the `odseq` package [@jehl2015], using the default threshold of 0.025.
The consensus of the remaining aligned sequences was assigned as the full-length ASV sequence.
Full-length ASV sequences with more than three ambiguous bases (i.e., no nucleotide >50% at a given position) were removed.
The count and sample distribution of reads assigned to each ASV were calculated in order to form a sample × ASV community matrix.
A similar process was used to generate a consensus ITS (ITS1--5.8S--ITS2) and LSU (LSU1--D1--LSU2--D2--LSU3--D3--LSU4) sequence for each ASV.
The process of assigning consensus full-length ASVs was carried out using the new `tzara` package for R, available on github (https://github.com/brendanf/tzara).

@@

Unique LSU regions from long amplicon ASVs were aligned using DECIPHER [@wright2015] with up to 10 iterations of progressive alignment and conserved RNA secondary structure calculation and 10 refinement iterations.
<!-- The LSU tree ends up not looking as nice as the tree which includes ITS (I don't know why, except that maybe 5.8S is more helpful than ITS is harmful?) so I'm using that. Maybe I don't need to mention the LSU-only tree.-->
This alignment was truncated at a position after the D3 region corresponding to base 907 of the *Saccharomyces cerevisiae* S288C reference sequence for LSU, because several sequences had introns
<!-- Just curious, what are Type 1 introns -AR--> <!-- It has to do with the enzyme that cuts them out, I think, but I don't actually know my introns are Type 1, just that this is a location where Type 1 introns are known to occur. -BF-->
after this position, as also observed in several fungal species by @holst-jensen1999.
Full length long amplicon ASVs (including ITS1, 5.8S, and ITS2 regions) were aligned and truncated in the same way.

@@

Taxonomic assignments of ASVs from the long amplicon dataset were refined using the phylogenetic tree (Supplementary Figure\ \@ref(fig:phylotax)).
A taxon at a particular rank was assigned to a node and all its descendents if that taxon was consistent with the reference-based taxonomic assignments for each of the descendents.
A taxon assignment was considered to be consistent if at least one algorithm assigned that taxon at greater than 50% confidence, or if no algorithm successfully classified the sequence at greater than 50% confidence.
The result of this process was to give a taxonomic assignment to ASVs which were previously unassigned if they were nested within a clade where which was consistently given an assignment, as well as clarify the assignment of ASVs where different algorithms had resulted in different assignments, but only one of these was consistent with the assignments of other ASVs in the same clade.
This refinement algorithm is referred to as "PHYLOTAX".

@@

Additionally, empirical distance-decay curves were generated by plotting mean community dissimilarity as a function of spatial distance, and fit to an exponential model of the form given by @legendre2012 using the `nls` function in R.
Points in the empirical distance-decay curve were weighted by the number of comparisons within the distance class and the inverse of the distance for the purposes of model fitting.
For datasets where the Mantel corellogram indicated spatial correlation between samples taken in seperate years, the model was re-fit with an additional term to represent temporal correlation:
$$D = C_0 + C_1\left[1 - \exp\left(-3  \left(\frac{d}{a_d} + \frac{t}{a_t}\right)\right)\right]$$
where $D$, $d$, and $t$ represent the community dissimilarity, spatial distance, and time lag between samples, respectively, and the parameters are $C_0$, the community dissimilarity from replicate samples ("nugget"); $C_0 + C_1$, the community dissimilarity at long distances ("sill"); $a_d$ the spatial range at which the community dissimilarity has moved 95% of the way from "nugget" to "sill", and $a_t$ the equivalent temporal range.
95% confidence intervals were calculated for the spatial and temporal range parameters by profiling using the `MASS` package in R.

@@

The number of sequencing reads and ASVs at each stage in the bioinformatics pipeline are shown in Table\ \@ref(tab:bioinfo).
Long (short) amplicon sequencing with PacBio yielded `r readd(rawcounts)["pb_500"]` (`r readd(rawcounts)["pb_483"]`) circular consensus reads, of which `r readd(demuxcounts)["pb_500"]` (`r readd(demuxcounts)["pb_483"]`) were successfully trimmed and demultiplexed.
Illumina sequencing yielded `r readd(rawcounts)["SH-2257"]` read pairs, of which `r readd(demuxcounts)["SH-2257"]` were successfully trimmed and demultiplexed, and `r readd(filtercounts_full)["SH-2257"]` passed quality filtering.
Ion Torrent sequencing yielded `r readd(demuxcounts)["is_057"]` demultiplexed reads.
`r readd(filtercounts_full)['pb_500']` (`r percent(readd(filtercounts_full)['pb_500']/readd(demuxcounts)['pb_500'], digits = 1, format = "f")`) of long amplicon PacBio reads passed length and quality filtering stages, compared with `r readd(filtercounts_full)['pb_483']` (`r percent(readd(filtercounts_full)['pb_483']/readd(demuxcounts)['pb_483'], digits = 1, format = "f")`) of short amplicon PacBio reads, `r readd(filtercounts_full)['SH-2257']` (`r percent(readd(filtercounts_full)['SH-2257']/readd(demuxcounts)['SH-2257'], digits = 1, format = "f")`) of Illumina read pairs, and `r readd(filtercounts_full)['is_057']` (`r percent(readd(filtercounts_full)['is_057']/readd(demuxcounts)['is_057'], digits = 1, format = "f")`) of Ion Torrent reads.

@@

## Region extraction
